Verification of Stencil Computations through Spatial Specifications

### 2016-07-07 paper.tex
space saving

\begin{figure}
%\vspace{-0.2em}
\begin{align*}
\setlength{\arraycolsep}{0.05em}
\begin{array}{c}
%\framebox{$\textit{region} \equiv \textit{region}'$} \\[0.7em]
\begin{array}{rll}
(\textsc{F\, \texttt{+} \,F}) \;\; &
\stenFwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
 \equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenFwdSR{m}{d}{s} \\[0.7em]
%
%
(\textsc{B\, \texttt{+} \,B}) \;\; &
\stenBwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
 \equiv \; & \stenBwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{m}{d}{s} \\[0.7em]
%
%
(\textsc{C\, \texttt{+} \,C}) \;\; &
\stenCenSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
\equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \; \stenCenSR{m}{d}{s} \\[0.7em]
%
%
(\textsc{C\, \texttt{+} \,F}) \;\; & \stenCenSR{n}{d}{r \vee s} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
                      \stenFwdSR{m}{d}{s} \\[0.7em]
%
%
(\textsc{C\, \texttt{+} \,B}) \;\; &
\stenCenSR{n}{d}{r \vee s} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
                      \stenBwdSR{m}{d}{s} \\[0.7em]
%
%
(\textsc{B\, \texttt{+} \,F}) \;\; &
\stenCenSR{n}{d}{r \vee s} \\
\equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{n}{d}{s}
  \\[0.7em]
%
%
(\textsc{R\, \texttt{+} \,F}) \;\; &
\stenFwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenFwdSR{n}{d}{r} \\[0.7em]
%
%
(\textsc{R\, \texttt{+} \,B}) \;\; &
\stenBwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenBwdSR{n}{d}{r} \\[0.7em]
%
%
(\textsc{R\, \texttt{+} \,C}) \;\; &
\stenCenS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenCenSR{n}{d}{r} \\[0.7em]
%
%
(\textsc{irrefl}) \;\; & (R \, \term{*} \, S^\irreflS) \, \term{+} \,
                         (R^\irreflS \, \term{*} \, S) \;
\equiv \; R \, \term{*} \, S \\[0.4em]
%
%
(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \\[0.4em]
%
(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
                       \texttt{+} \; S \\[0.4em]
%
(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.4em]
(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
                       \texttt{*} \; S \\[0.4em]
%
(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.4em]
(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T)
\end{array}
%
%\framebox{${\textit{specDec} :: \texttt{v}} \equiv
%{\textit{specDec'} :: \texttt{v}}$} \\[0.7em]
%\begin{array}{rl}
% MUTUAL
%(\textsc{mutual}) \; &
%\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
%  :: \texttt{w}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
%  \texttt{v} \\[0.5em]
%(\textsc{coalesce}) \; &
%\texttt{stencil} \; \texttt{dependency} (\bar{v})
%  :: \texttt{v} \\
%& ; \texttt{stencil} \; \texttt{dependency} (\bar{w})
%  :: \texttt{v}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
%  \texttt{v} \\[0.5em]
%\end{array}
\end{array}
\end{align*}
\caption{Equations on specifications (regions)}
\vspace{-1em}
\label{fig:equations}
\end{figure}

@@

\begin{definition} The $\mathsf{toSpec}$ constructor is defined:

%\newcommand{\llower}{\vect{x}^L_i}
%\newcommand{\lupper}{\vect{x}^U_i}
\newcommand{\llower}{l}
\newcommand{\lupper}{u}
{\scalebox{0.97}{
\hspace{-1.8em}
\begin{minipage}{1\linewidth}
\begin{align*}
&\mathsf{toSpec} \; \llower \; \lupper = \\
&\hspace{-0.4em}\begin{cases}
%% FWD
\exact{\stenFwdS{\lupper}{i}}  & \llower = 0 \, \wedge \, \lupper > 0 \\[0.2em]
\exact{\stenFwdSR{\lupper}{i}{\irreflS}}\!\!\!\! & \llower = 1 \, \wedge \, \lupper > 0 \\[0.2em]
%% BWD
\exact{\stenBwdS{|\llower|}{i}}  & \llower < 0 \, \wedge \, \lupper = 0 \\[0.2em]
\exact{\stenBwdSR{|\llower|}{i}{\irreflS}}\!\!\!\! & \llower < 0 \, \wedge \, \lupper = -1 \\[0.2em]
%% CEN
\exact{\stenCenS{\lupper}{i}}& \llower < 0 \wedge \lupper > 0 \wedge |\llower| = \lupper \\[0.2em]
\exactR(\stenBwdS{|\llower|}{i} & \llower < 0 \wedge \lupper > 0  \wedge |\llower| \neq \lupper \\
\qquad\qquad \texttt{+} \, \stenFwdS{\lupper}{i}) & \\[0.2em]
% BOUND
\upperB{\stenFwdS{\lupper}{i}} & \llower > 1 \\[0.2em]
\upperB{\stenBwdS{|\llower|}{i}} \quad & \llower < (-1)
 \\[0.2em]
% ABSOLUTE REP
\exact{\epsilon} & \llower = \infty \, \vee \, \lupper = \infty
  \end{cases}
\end{align*}
\end{minipage}
}}\\

@@

\begin{lemma}[$\hat{\odot}$, $\hat{\oplus}$ soundness] $\forall S,
  T \in \specDNF$ where $S \neq \epsilon \wedge T \neq \epsilon$:
\begin{align*}
\interp{S \; \hat{\odot} \; T}_n = \interp{S}_n \otimes \interp{T}_n
  \qquad
\interp{S \; \hat{\oplus} \; T}_n = \interp{S}_n \cup \interp{T}_n
\end{align*}
\label{lem:alg-soundness}
\vspace{-2em}
\end{lemma}
%
\noindent
Thus our model validates the specification synthesis process which concludes the
inference procedure. For our example, we infer and then synthesise the
5-point specification:
%%
\begin{minted}[linenos=false]{fortran}
!= stencil centered(depth=1, dim=1)*reflexive(dim=2) + centered(depth=1, dim=2)*reflexive(dim=1):: a
\end{minted}
%%

@@

In checking mode, the access pattern of a stencil computation in the
source language is verified against a specification. The model of
Section~\ref{sec:semantics} again aids this process.  Checking
proceeds by generating a model $M$ from a specification and generating a set $N$ of indexing
schemes from the source code. The source code is then valid with
respect to the specification if $N$ is \emph{consistent} with $M$. The
notion of consistency is built out of index scheme 
\emph{generalisation} (Sect.~\ref{subsec:schemes-and-gen}). 
Consistency is defined in Figure~\ref{fig:consistency},
but we first set up some intermediate definitions before unpacking its definition.

@@

Package   & {Total} & {Parsed} & {Total} & {Parsed}  \\ % braces prevent headers from being formatted as numbers
\hline
UM        & \umFiles                       & \umparseOk                      & \umLoC                      & \umlinesParsed \\
E3MG      & \ethreemgeaFiles               & \ethreemgeaparseOk              & \ethreemgeaLoC              & \ethreemgealinesParsed \\
BLAS      & \blasFiles                     & \blasparseOk                    & \blasLoC                    & \blaslinesParsed \\
Hybrid4   & \hybridfourFiles               & \hybridfourparseOk              & \hybridfourLoC              & \hybridfourlinesParsed \\
GEOS-Chem & \geoschemFiles                 & \geoschemparseOk                & \geoschemLoC                & \geoschemlinesParsed \\
Navier    & \navierFiles                   & \navierparseOk                  & \navierLoC                  & \navierlinesParsed \\
CP        & \computationalphysicstwoFiles  & \computationalphysicstwoparseOk & \computationalphysicstwoLoC & \computationalphysicstwolinesParsed \\
\hline
\textbf{Total} & \overallFiles  & \overallparseOk & \overallLoC & \overalllinesParsed \\
\hline
\end{tabular}
\caption{Summary of software packages used for
  evaluation\label{tab:corpus}}
\vspace{-1em}
\end{table}

@@

\begin{table}
\sisetup{table-format=4,table-text-alignment=right}
\footnotesize
\centering
\begin{tabular}{lS@{~~~}rS@{~~~}rSS[table-format=1]}
\hline
  & \multicolumn{2}{c}{Potential} & \multicolumn{2}{c}{Actual} & {Exact} & {Inexact}    \\
{Package}     & {\#} & \si{\percent} & {\#} & \si{\percent} \\
\hline
UM            & \umtickAssign & \SI{\umtickAssignPercent}{\percent} &
                \umtickAssignSuccess & \SI{\umtickAssignSuccessPercent}{\percent} &
                \umnumStencilLines &
                \umboundedBoth  \\
E3MG          & \ethreemgeatickAssign & \SI{\ethreemgeatickAssignPercent}{\percent} &
                \ethreemgeatickAssignSuccess & \SI{\ethreemgeatickAssignSuccessPercent}{\percent} &
                \ethreemgeanumStencilLines &
                \ethreemgeaboundedBoth \\
BLAS          & \blastickAssign & \SI{\blastickAssignPercent}{\percent} &
                \blastickAssignSuccess & \SI{\blastickAssignSuccessPercent}{\percent} &
                \blasnumStencilLines &
                \blasboundedBoth \\
Hybrid4       & \hybridfourtickAssign & \SI{\hybridfourtickAssignPercent}{\percent} &
                \hybridfourtickAssignSuccess & \SI{\hybridfourtickAssignSuccessPercent}{\percent} &
                \hybridfournumStencilLines &
                \hybridfourboundedBoth\\
GEOS-Chem     & \geoschemtickAssign & \SI{\geoschemtickAssignPercent}{\percent} &
                \geoschemtickAssignSuccess & \SI{\geoschemtickAssignSuccessPercent}{\percent} &
                \geoschemnumStencilLines &
                \geoschemboundedBoth \\
Navier        & \naviertickAssign & \SI{\naviertickAssignPercent}{\percent} &
                \naviertickAssignSuccess & \SI{\naviertickAssignPercent}{\percent} &
                \naviernumStencilLines &
                \navierboundedBoth \\
CP            & \computationalphysicstwotickAssign & \SI{\computationalphysicstwotickAssignPercent}{\percent} &
                \computationalphysicstwotickAssignSuccess & \SI{\computationalphysicstwotickAssignSuccessPercent}{\percent} &
                \computationalphysicstwonumStencilLines &
                \computationalphysicstwoboundedBoth \\
\hline
\textbf{Total}& \overalltickAssign & \SI{\overalltickAssignPercent}{\percent}  &
                \overalltickAssignSuccess & \SI{\overalltickAssignSuccessPercent}{\percent} &
                \overallnumStencilLines &
                \overallboundedBoth    \\
\hline
\end{tabular}
\caption{Stencil assignment rates\label{tab:tickAssign}}
\vspace{-0.8em}
\end{table}

@@

One relatively basic example of a stencil computation is two-dimensional
Jacobi iteration that repeatedly goes through each cell in a matrix
and computes the average value of the four adjacent cells. The kernel
can be expressed in a line of Fortran code:
\begin{minted}{fortran}
  a(i,j) = (a(i-1,j)+a(i+1,j)+a(i,j+1)+a(i,j-1))/4
\end{minted}
and CamFort is able to infer a precise specification for this:
\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
  != stencil reflexive(dim=1)*centered(depth=1,dim=2,irreflexive)+ reflexive(dim=2)*centered(depth=1,dim=1,irreflexive) :: a
\end{minted}
We then tested the error-detecting ability of our approach by 
generating a family of 6,561 Jacobi iteration programs by perturbing
the indices used in the following manner: each access of the array was
templated as \mintinline{fortran}{a(i+ki,j+kj)} and then all possible
combinations of \mintinline{fortran}{ki} and \mintinline{fortran}{kj}
were drawn from the set $\{-1,0,1\}$. There are four uses of
\mintinline{fortran}{ki} and four of \mintinline{fortran}{kj}, hence
$3^8=6561$ possible perturbations of this kind. This particular setup
tests the class of errors in which the programmer typed an incorrect
number but at least got the ordering of the indices correct, although
CamFort would also detect a mixup of variables in this case as well.

@@

\citet{kamil2016verified} propose a technique they call
\emph{verified lifting} for mapping low-level, possibly
greatly optimised, stencil computations into a high-level
predicate language using inductive program synthesis.
The aim is that these high-level abstract specifications can
then be translated into high-performance DSLs, easing the
use of high-performance DSLs. Compared to our approach,
their specification language is more fine grained, capturing
the exact indexing pattern of a stencil. For example, the following
is an example post-condition in their language, generated by their
compiler \textsc{stng} from a Fortran stencil computation~\cite [p.3]{kamil2016verified}:
%
\begin{align*}
& \textit{post}(\texttt{a}, \texttt{b}) \equiv \forall \texttt{imin}+1
\leq \texttt{i} \leq \texttt{imax}, \texttt{jmin} \leq \texttt{j} \leq
\texttt{jmax} \\
& \qquad \texttt{a}(\texttt{i},\texttt{j}) =
\texttt{b}(\texttt{i}-1,\texttt{j}) + \texttt{b}(\texttt{i},\texttt{j})
\end{align*}
%
Notably, the bounds of the stencil and the exact indexing pattern
are captured. Our approach aims to be much more abstract, to
facilitate human reasoning, and avoid low-level lexical
errors. However, \textsf{stng} specifications are generated, \ie{},
are never written by a programmer, avoiding this
problem. Their inductive program synthesis approach allows
specifications to extracted even from optimised and unrolled code. For
example, the above post condition is generated from the code~\cite [p.3]{kamil2016verified}:
%
\begin{minted}{fortran}
do j=jmin, jmax
    t = b(imin, j)
    do i=imin+1,imax
        q = b(i, j)
        a(i, j) = q + t
        t = q
    end do
end do
\end{minted}
%
Thus, \textsc{stng} can understand cross-loop data-dependencies
in order to assign the above post-condition. CamFort is not currently
able to ascribe a specification in the above case as the use of
\texttt{t} on line 5 maps only to the definition on line 2, and no on
line 6. Extending our approach to follow loop-carried dependencies is
future work. An alternate approach for us would be to build on top of
the work of Kamil~\emph{et al.} by inferring our specifications from
\textsc{stng} (as an intermediate step).

>>>>

\begin{figure}
%\vspace{-0.2em}
\begin{align*}
\setlength{\arraycolsep}{0.05em}
\begin{array}{c}
%\framebox{$\textit{region} \equiv \textit{region}'$} \\[0.6em]
\begin{array}{rll}
(\textsc{F\, \texttt{+} \,F}) \;\; &
\stenFwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
 \equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenFwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{B\, \texttt{+} \,B}) \;\; &
\stenBwdSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
 \equiv \; & \stenBwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,C}) \;\; &
\stenCenSR{n \, \textsf{max} \, m}{d}{r \vee s} \\
\equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \; \stenCenSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,F}) \;\; & \stenCenSR{n}{d}{r \vee s} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
                      \stenFwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{C\, \texttt{+} \,B}) \;\; &
\stenCenSR{n}{d}{r \vee s} \\
\textit{$m \leq n$} \; \equiv \; & \stenCenSR{n}{d}{r} \; \texttt{+} \;
                      \stenBwdSR{m}{d}{s} \\[0.6em]
%
%
(\textsc{B\, \texttt{+} \,F}) \;\; &
\stenCenSR{n}{d}{r \vee s} \\
\equiv \; & \stenFwdSR{n}{d}{r} \; \texttt{+} \; \stenBwdSR{n}{d}{s}
  \\[0.6em]
%
%
(\textsc{R\, \texttt{+} \,F}) \;\; &
\stenFwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenFwdSR{n}{d}{r} \\[0.6em]
%
%
(\textsc{R\, \texttt{+} \,B}) \;\; &
\stenBwdS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenBwdSR{n}{d}{r} \\[0.6em]
%
%
(\textsc{R\, \texttt{+} \,C}) \;\; &
\stenCenS{n}{d} \\
\equiv \; & \stenReflSR{d} \; \texttt{+} \; \stenCenSR{n}{d}{r} \\[0.6em]
%
%
(\textsc{irrefl}) \;\; & (R \, \term{*} \, S^\irreflS) \, \term{+} \,
                         (R^\irreflS \, \term{*} \, S) \;
\equiv \; R \, \term{*} \, S \\[0.4em]
%
%
(\texttt{+}\textsc{idem}) \;\; & S \; \texttt{+} \; S \; \equiv \; S \\[0.4em]
%
(\texttt{+}\textsc{comm}) \;\; & S \; \texttt{+} \; T \; \equiv \; T \;
                       \texttt{+} \; S \\[0.4em]
%
(\texttt{+}\textsc{assoc}) \;\; & R \; \texttt{+} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{+} \; S) \; \texttt{+} \; T \\[0.4em]
(\texttt{*}\textsc{comm}) \;\; & S \; \texttt{*} \; T \; \equiv \; T \;
                       \texttt{*} \; S \\[0.4em]
%
(\texttt{*}\textsc{assoc}) \;\; & R \; \texttt{*} \; (S \; \texttt{*} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{*} \; T \\[0.4em]
(\textsc{dist}) \;\; & R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T)
\end{array}
%
%\framebox{${\textit{specDec} :: \texttt{v}} \equiv
%{\textit{specDec'} :: \texttt{v}}$} \\[0.7em]
%\begin{array}{rl}
% MUTUAL
%(\textsc{mutual}) \; &
%\texttt{stencil} \; \texttt{dependency} (\texttt{v}), \texttt{mutual}
%  :: \texttt{w}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\texttt{w}), \texttt{mutual} ::
%  \texttt{v} \\[0.5em]
%(\textsc{coalesce}) \; &
%\texttt{stencil} \; \texttt{dependency} (\bar{v})
%  :: \texttt{v} \\
%& ; \texttt{stencil} \; \texttt{dependency} (\bar{w})
%  :: \texttt{v}
%\\
%\equiv \; & \texttt{stencil} \; \texttt{dependency} (\bar{v}, \bar{w}) ::
%  \texttt{v} \\[0.5em]
%\end{array}
\end{array}
\end{align*}
\caption{Equations on specifications (\emph{regions})}
\vspace{-1em}
\label{fig:equations}
\end{figure}

@@

\begin{definition} The $\mathsf{toSpec}$ constructor is defined
in Figure~\ref{fig:to-spec}

\begin{figure}[t]
%\newcommand{\llower}{\vect{x}^L_i}
%\newcommand{\lupper}{\vect{x}^U_i}
\vspace{-0.7em}
\newcommand{\llower}{l}
\newcommand{\lupper}{u}
{\scalebox{0.97}{
\begin{minipage}{1\linewidth}
\begin{align*}
&\mathsf{toSpec} \; \llower \; \lupper = \\
&\hspace{-0.4em}\begin{cases}
%% FWD
\exact{\stenFwdS{\lupper}{i}}  & \llower = 0 \, \wedge \, \lupper > 0 \\[0.2em]
\exact{\stenFwdSR{\lupper}{i}{\irreflS}}\!\!\!\! & \llower = 1 \, \wedge \, \lupper > 0 \\[0.2em]
%% BWD
\exact{\stenBwdS{|\llower|}{i}}  & \llower < 0 \, \wedge \, \lupper = 0 \\[0.2em]
\exact{\stenBwdSR{|\llower|}{i}{\irreflS}}\!\!\!\! & \llower < 0 \, \wedge \, \lupper = -1 \\[0.2em]
%% CEN
\exact{\stenCenS{\lupper}{i}}& \llower < 0 \wedge \lupper > 0 \wedge |\llower| = \lupper \\[0.2em]
\exactR(\stenBwdS{|\llower|}{i} & \llower < 0 \wedge \lupper > 0  \wedge |\llower| \neq \lupper \\
\qquad\qquad \texttt{+} \, \stenFwdS{\lupper}{i}) & \\[0.2em]
% BOUND
\upperB{\stenFwdS{\lupper}{i}} & \llower > 1 \\[0.2em]
\upperB{\stenBwdS{|\llower|}{i}} \quad & \llower < (-1)
 \\[0.2em]
% ABSOLUTE REP
\exact{\epsilon} & \llower = \infty \, \vee \, \lupper = \infty
  \end{cases}
\end{align*}
\end{minipage}
}}
\caption{\textsf{toSpec} constructor of $\mathsf{A}(\specDNF)$ values.}
\label{fig:tospec}
\vspace{-0.9em}
\end{figure}

@@

\begin{lemma}[$\hat{\odot}$, $\hat{\oplus}$ soundness]$\forall S,
  T \in \specDNF$ where $S\!\neq \epsilon \wedge T\!\neq \epsilon$:
\begin{align*}
\interp{S \; \hat{\odot} \; T}_n = \interp{S}_n \otimes \interp{T}_n
  \qquad
\interp{S \; \hat{\oplus} \; T}_n = \interp{S}_n \cup \interp{T}_n
\end{align*}
\label{lem:alg-soundness}
\vspace{-2em}
\end{lemma}
%
\noindent
Thus our model validates the specification synthesis process which concludes the
inference procedure. For our example, we infer and then synthesise the
5-point specification:
%%
\begin{minted}[linenos=false]{fortran}
!= stencil centered(depth=1, dim=1)*reflexive(dim=2) + centered(depth=1, dim=2)*reflexive(dim=1):: a
\end{minted}
%%

@@

In checking mode, the access pattern of a stencil computation in the
source language is verified against a specification. The model of
Section~\ref{sec:semantics} again aids this process.  Checking
proceeds by generating a model $M$ from a specification and generating a set $N$ of indexing
schemes from the source code. The source code is then valid with
respect to the specification if $N$ is \emph{consistent} with $M$. The
notion of consistency is built out of index scheme 
\emph{generalisation} (Sect.~\ref{subsec:schemes-and-gen}). 
%Consistency is defined in Figure~\ref{fig:consistency},
We first set up some intermediate definitions.% before unpacking its definition.

@@

Package   & {Total} & {Parsed} & {Total} & {Parsed}  \\ % braces prevent headers from being formatted as numbers
\hline
UM        & \umFiles                       & \umparseOk                      & \umLoC                      & \umlinesParsed \\
E3MG      & \ethreemgeaFiles               & \ethreemgeaparseOk              & \ethreemgeaLoC              & \ethreemgealinesParsed \\
BLAS      & \blasFiles                     & \blasparseOk                    & \blasLoC                    & \blaslinesParsed \\
Hybrid4   & \hybridfourFiles               & \hybridfourparseOk              & \hybridfourLoC              & \hybridfourlinesParsed \\
GEOS-Chem & \geoschemFiles                 & \geoschemparseOk                & \geoschemLoC                & \geoschemlinesParsed \\
Navier    & \navierFiles                   & \navierparseOk                  & \navierLoC                  & \navierlinesParsed \\
CP        & \computationalphysicstwoFiles  & \computationalphysicstwoparseOk & \computationalphysicstwoLoC & \computationalphysicstwolinesParsed \\
\hline
\textbf{Total} & \overallFiles  & \overallparseOk & \overallLoC & \overalllinesParsed \\
\hline
\end{tabular}
\caption{Summary of software packages used for
  evaluation\label{tab:corpus}}
\vspace{-0.7em}
\end{table}

@@

\begin{table}
\sisetup{table-format=4,table-text-alignment=right}
\footnotesize
\centering
\begin{tabular}{lS@{~~~}rS@{~~~}rSS[table-format=1]}
\hline
  & \multicolumn{2}{c}{Potential} & \multicolumn{2}{c}{Actual} & {Exact} & {Inexact}    \\
{Package}     & {\#} & \si{\percent} & {\#} & \si{\percent} \\
\hline
UM            & \umtickAssign & \SI{\umtickAssignPercent}{\percent} &
                \umtickAssignSuccess & \SI{\umtickAssignSuccessPercent}{\percent} &
                \umnumStencilLines &
                \umboundedBoth  \\
E3MG          & \ethreemgeatickAssign & \SI{\ethreemgeatickAssignPercent}{\percent} &
                \ethreemgeatickAssignSuccess & \SI{\ethreemgeatickAssignSuccessPercent}{\percent} &
                \ethreemgeanumStencilLines &
                \ethreemgeaboundedBoth \\
BLAS          & \blastickAssign & \SI{\blastickAssignPercent}{\percent} &
                \blastickAssignSuccess & \SI{\blastickAssignSuccessPercent}{\percent} &
                \blasnumStencilLines &
                \blasboundedBoth \\
Hybrid4       & \hybridfourtickAssign & \SI{\hybridfourtickAssignPercent}{\percent} &
                \hybridfourtickAssignSuccess & \SI{\hybridfourtickAssignSuccessPercent}{\percent} &
                \hybridfournumStencilLines &
                \hybridfourboundedBoth\\
GEOS-Chem     & \geoschemtickAssign & \SI{\geoschemtickAssignPercent}{\percent} &
                \geoschemtickAssignSuccess & \SI{\geoschemtickAssignSuccessPercent}{\percent} &
                \geoschemnumStencilLines &
                \geoschemboundedBoth \\
Navier        & \naviertickAssign & \SI{\naviertickAssignPercent}{\percent} &
                \naviertickAssignSuccess & \SI{\naviertickAssignPercent}{\percent} &
                \naviernumStencilLines &
                \navierboundedBoth \\
CP            & \computationalphysicstwotickAssign & \SI{\computationalphysicstwotickAssignPercent}{\percent} &
                \computationalphysicstwotickAssignSuccess & \SI{\computationalphysicstwotickAssignSuccessPercent}{\percent} &
                \computationalphysicstwonumStencilLines &
                \computationalphysicstwoboundedBoth \\
\hline
\textbf{Total}& \overalltickAssign & \SI{\overalltickAssignPercent}{\percent}  &
                \overalltickAssignSuccess & \SI{\overalltickAssignSuccessPercent}{\percent} &
                \overallnumStencilLines &
                \overallboundedBoth    \\
\hline
\end{tabular}
\caption{Stencil specification assignment rates\label{tab:tickAssign}}
\vspace{-0.8em}
\end{table}

@@

One relatively basic example of a stencil computation is two-dimensional
Jacobi iteration that repeatedly goes through each cell in a matrix
and computes the average value of the four adjacent cells. The kernel
can be expressed in a line of Fortran code:
\begin{minted}{fortran}
  a(i,j) = (a(i-1,j)+a(i+1,j)+a(i,j+1)+a(i,j-1))/4
\end{minted}
and CamFort is able to infer a precise specification for this:
\begin{minted}[breakafter=+:,breakindent=-0.6em,breaksymbolsep=0.4em,linenos=false,xleftmargin=-0.5em]{fortran}
  != stencil reflexive(dim=1)*centered(depth=1,dim=2,irreflexive)+reflexive(dim=2)*centered(depth=1,dim=1,irreflexive)
\end{minted}
assigned to \term{a}. 
We then tested the error-detecting ability of our approach by 
generating a family of 6,561 Jacobi iteration programs by perturbing
the indices used in the following manner: each access of the array was
templated as \mintinline{fortran}{a(i+ki,j+kj)} and then all possible
combinations of \mintinline{fortran}{ki} and \mintinline{fortran}{kj}
were drawn from the set $\{-1,0,1\}$. There are four uses of
\mintinline{fortran}{ki} and four of \mintinline{fortran}{kj}, hence
$3^8=6561$ possible perturbations of this kind. This particular setup
tests the class of errors in which the programmer typed an incorrect
number but at least got the ordering of the indices correct, although
CamFort would also detect a mixup of variables in this case as well.

@@

\citet{kamil2016verified} propose a technique they call
\emph{verified lifting} for mapping low-level, possibly
greatly optimised, stencil computations into a high-level
predicate language using inductive program synthesis.
The aim is that the high-level abstract specifications can
be automatically translated into high-performance DSLs. Compared to our approach,
their specification language is more fine grained, capturing
the exact indexing pattern of a stencil. For example, the following
is a post-condition generated by their
compiler \textsc{stng} from a Fortran stencil computation~\cite [p.3]{kamil2016verified}:
%
\begin{align*}
& \textit{post}(\texttt{a}, \texttt{b}) \equiv \forall \texttt{imin}+1
\leq \texttt{i} \leq \texttt{imax}, \texttt{jmin} \leq \texttt{j} \leq
\texttt{jmax} \\
& \qquad \texttt{a}(\texttt{i},\texttt{j}) =
\texttt{b}(\texttt{i}-1,\texttt{j}) + \texttt{b}(\texttt{i},\texttt{j})
\end{align*}
%
Notably, the bounds of the stencil and the exact indexing pattern
are captured. Our approach aims to be much more abstract, to
facilitate human reasoning, and avoid low-level lexical
errors. However, \textsc{stng} specifications are generated, \ie{},
are never written by a programmer, avoiding this
problem. Their inductive program synthesis approach allows
specifications to extracted even from optimised and unrolled code. For
example, the above post condition is generated from the code~\cite [p.3]{kamil2016verified}:
%
\begin{minted}{fortran}
do j=jmin, jmax
    t = b(imin, j)
    do i=imin+1,imax
        q = b(i, j)
        a(i, j) = q + t
        t = q
    end do
end do
\end{minted}
%
Thus, \textsc{stng} can understand cross-loop data-dependencies
in order to assign the above post-condition. CamFort is not currently
able to ascribe a specification in the above case as the use of
\texttt{t} on line 5 maps only to the definition on line 2, and not to
that on line 6. Extending our approach to follow loop-carried dependencies is
future work. An alternate approach for us would be to build on top of
the work of Kamil~\emph{et al.} by inferring our specifications from
\textsc{stng} (as an intermediate step).
